---
title: "HOMEWORK 2"
author: "ilaydacelenk"
date: "27/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, conflict=FALSE, warning=FALSE, message=FALSE, error = FALSE, comment = FALSE)
```

```{r libraries}
#importing the libraries
library(scatterplot3d) # part(a)
library(tidyverse) #filter etc.
library(ggplot2) #plot in reduced dimensions
library(reshape) #melt
library(directlabels)
library(dplyr)
library(DT) #datatable
```

### Introduction:
Given uWaveGestureLibrary, consisting over 4000 instances from 8 people with 8 different gestures, our aim is to visualize the gestures. Then, we will try to apply dimensionality reduction. There are 8 different gesture classes for each axis, x, y and z.

### Tasks:


#### A. Read the Data and Plot an Instance from Each Gesture
```{r read-data}
#reading the data from repository
x_train=read.table("https://github.com/BU-IE-582/fall20-ilaydacelenk/blob/master/files/HW2_data/uWaveGestureLibrary_X_TRAIN.txt?raw=true")
y_train=read.table("https://github.com/BU-IE-582/fall20-ilaydacelenk/blob/master/files/HW2_data/uWaveGestureLibrary_Y_TRAIN.txt?raw=true")
z_train=read.table("https://github.com/BU-IE-582/fall20-ilaydacelenk/blob/master/files/HW2_data/uWaveGestureLibrary_Z_TRAIN.txt?raw=true")
```

For the x-axis, let's find the indices where we see each gesture for the first time. It is enough to look at x_train since y_train and z_train will have the same indices for the first occurrences of each gesture.

In order to find the velocity vector, we need to compute the cumulative sum of acceleration over time. Since we got the indices, it is easy to sum the 

```{r 3D-scatter-plot}
#the first instances when we see gesture of class i, i from 1 to 8
index_vector = c()
for(i in 1:8) 
{
  index_vector <- index_vector %>% append(match(i, x_train$V1)) 
}

# Creating cumulative sum matrices 2 times to get first the velocity vector and then the location vector for each instance
# using velocity(t)=cumsum(acceleration)*change in t
# using position(t)=cumsum(velocity)*change in t

loc_matrix <- function(acc_matrix){
  order_of_instances <- acc_matrix$V1
  acc_matrix$V1 <- 0
  loc <- acc_matrix %>% mutate(V1=0) %>% apply(1, cumsum) %>% t() %>% apply(1, cumsum) %>% t() %>% as.data.frame() %>% mutate(V1=order_of_instances)
  return(loc)
}
  
x_position <- loc_matrix(x_train) %>% as.matrix()
y_position <- loc_matrix(y_train) %>% as.matrix()
z_position <- loc_matrix(z_train) %>% as.matrix()


plot_3d <- function(indices){
  for(i in 1:8) {
    scatterplot3d(x_position[index_vector[i],-1], y_position[index_vector[i],-1], z_position[index_vector[i],-1], main = paste("Gesture Class ", x_position[index_vector[i],1]) ,xlab = "X Axis", ylab = "Y Axis", zlab = "Z Axis", col.grid="lightblue",type = "p", color = "red")
  }
}
  
plot_3d(index_vector)


```




#### B. Dimensionality Reduction - PCA on the Whole Data

Here we are dealing with multivariate time series and we would like to reduce it to a univariate time series. In order to achieve this, we transform the data into the long format first. 

Column V2 is time 1 and column V316 represents the time index 315. 

```{r long-format}

x_long <- x_position %>% as.data.frame() %>% mutate(id = seq.int(nrow(x_position))) %>% melt(id.vars = c("id", "V1")) %>% transmute(time_series_id=id, time_index=variable, X=value, class=V1)

y_long <- y_position %>% as.data.frame() %>% mutate(id = seq.int(nrow(x_position))) %>% melt(id.vars = c("id", "V1")) %>% transmute(time_series_id=id, time_index=variable, Y=value, class=V1)

z_long <- z_position %>% as.data.frame() %>% mutate(id = seq.int(nrow(x_position))) %>% melt(id.vars = c("id", "V1")) %>% transmute(time_series_id=id, time_index=variable, Z=value, class=V1)

xyz_long <- x_long %>% select(-class) %>% mutate(Y=y_long$Y, Z=z_long$Z, class = x_long$class)  %>% mutate(time_index = sub(".", "", time_index)) 

xyz_long$time_index <- as.numeric(xyz_long$time_index) - 1 

```

Then, we apply PCA on X, Y and Z. This will be applied to the whole data. 
According to the PCA results, the variance covered by the first component is approximately 1.27 and it is the linear combination 0.209*X + 0.723*Y + 0.658*Z. Looking at the proportions Comp.1 covers 54% of the variance, Comp.2 covers 36% and Comp.3 covers 10% and they all add up to 100% of the variance. Since we would like to turn it into a univariate time series, we will only choose Comp.1 which has the highest percentage of variance covering.

```{r reduce-dims}
pca <- princomp(xyz_long[,3:5], cor=T) #scale differences T
print(summary(pca, loadings=T))
```


```{r plot-2-from-each}

xyz_long_uni <- xyz_long %>% mutate(value = predict(pca,xyz_long[,3:5])[,1]) %>% select(-X, -Y, -Z)

ex_class <- xyz_long_uni %>% filter(time_series_id %in% c(11, 17, 15, 20, 4, 13, 5, 8, 2, 3, 1, 10, 7, 12, 6, 21)) %>% mutate(time_series_id = as_factor(time_series_id), class = as_factor(class))

ggplot(ex_class, aes(x=time_index, y=value, colour=time_series_id)) + geom_line() + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + scale_color_discrete(name = "Classes", labels = c("Class6", "Class5", "Class5", "Class3", "Class4", "Class8", "Class7", "Class4", "Class6", "Class1", "Class7", "Class3", "Class2", "Class1", "Class2", "Class8")) + geom_dl(aes(label = class), method = list(dl.combine("last.points")), cex = 0.8) 

ggplot(ex_class, aes(x=time_index, y=value, colour=class)) + geom_line() + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + scale_color_discrete(name = "Classes")

```

#### C. Dimensionality Reduction - PCA on the Data from Each Gesture


Using the long format for the position data, we apply PCA on X, Y and Z values for each gesture. This can be achieved by filtering the data in long format. 

According to the PCA results, with respect to the order of the gestures, the proportions of variance covered by the first components are approximately 0.56, 0.42, 0.50, 0.55, 0.66, 0.60, 0.56 and 0.50. 

Since this analysis made on each gesture separately, the linear combinations of X, Y and Z coordinates have different coefficients. For example, for gesture 8, we have 0.664*X+0.337*Y+0.668*Z. With the PCA applied on the whole data, we had 0.209*X + 0.723*Y + 0.658*Z. The coefficients of X coordinates have a huge difference. 

Generally, same gestures follow similar graphs. But when we plot them all together, it is observed that different gestures also follow a similar pattern. This shows that this may not be a very good method to distinguish gestures. 

```{r pca}
xyz_long_1 <- xyz_long %>% filter(class==1)
pca1 <- princomp(xyz_long_1[,3:5], cor=T) #scale differences T
print(summary(pca1, loadings=T))
xyz_long_uni_1 <- xyz_long_1 %>% mutate(value = predict(pca1,xyz_long_1[,3:5])[,1]) %>% select(-X, -Y, -Z) %>% filter(time_series_id==11 | time_series_id==17) %>% mutate(time_series_id = as_factor(time_series_id))
ggplot(xyz_long_uni_1) + geom_line(aes(x=time_index, y=value, colour=time_series_id)) + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("Time Series from Class 1")

xyz_long_2 <- xyz_long %>% filter(class==2)
pca2 <- princomp(xyz_long_2[,3:5], cor=T) #scale differences T
print(summary(pca2, loadings=T))
xyz_long_uni_2 <- xyz_long_2 %>% mutate(value = predict(pca2,xyz_long_2[,3:5])[,1]) %>% select(-X, -Y, -Z) %>% filter(time_series_id==15 | time_series_id==20) %>% mutate(time_series_id = as_factor(time_series_id))
ggplot(xyz_long_uni_2) + geom_line(aes(x=time_index, y=value, colour=time_series_id)) + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("Time Series from Class 2")

xyz_long_3 <- xyz_long %>% filter(class==3)
pca3 <- princomp(xyz_long_3[,3:5], cor=T) #scale differences T
print(summary(pca3, loadings=T))
xyz_long_uni_3 <- xyz_long_3 %>% mutate(value = predict(pca3,xyz_long_3[,3:5])[,1]) %>% select(-X, -Y, -Z) %>% filter(time_series_id==4 | time_series_id==13) %>% mutate(time_series_id = as_factor(time_series_id))
ggplot(xyz_long_uni_3) + geom_line(aes(x=time_index, y=value, colour=time_series_id)) + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("Time Series from Class 3")

xyz_long_4 <- xyz_long %>% filter(class==4)
pca4 <- princomp(xyz_long_4[,3:5], cor=T) #scale differences T
print(summary(pca4, loadings=T))
xyz_long_uni_4 <- xyz_long_4 %>% mutate(value = predict(pca4,xyz_long_4[,3:5])[,1]) %>% select(-X, -Y, -Z) %>% filter(time_series_id==5 | time_series_id==8) %>% mutate(time_series_id = as_factor(time_series_id))
ggplot(xyz_long_uni_4) + geom_line(aes(x=time_index, y=value, colour=time_series_id)) + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("Time Series from Class 4")

xyz_long_5 <- xyz_long %>% filter(class==5)
pca5 <- princomp(xyz_long_5[,3:5], cor=T) #scale differences T
print(summary(pca5, loadings=T))
xyz_long_uni_5 <- xyz_long_5 %>% mutate(value = predict(pca5,xyz_long_5[,3:5])[,1]) %>% select(-X, -Y, -Z) %>% filter(time_series_id==2 | time_series_id==3) %>% mutate(time_series_id = as_factor(time_series_id))
ggplot(xyz_long_uni_5) + geom_line(aes(x=time_index, y=value, colour=time_series_id)) + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("Time Series from Class 5")

xyz_long_6 <- xyz_long %>% filter(class==6)
pca6 <- princomp(xyz_long_6[,3:5], cor=T) #scale differences T
print(summary(pca6, loadings=T))
xyz_long_uni_6 <- xyz_long_6 %>% mutate(value = predict(pca6,xyz_long_6[,3:5])[,1]) %>% select(-X, -Y, -Z) %>% filter(time_series_id==1 | time_series_id==10) %>% mutate(time_series_id = as_factor(time_series_id))
ggplot(xyz_long_uni_6) + geom_line(aes(x=time_index, y=value, colour=time_series_id)) + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("Time Series from Class 6")

xyz_long_7 <- xyz_long %>% filter(class==7)
pca7 <- princomp(xyz_long_7[,3:5], cor=T) #scale differences T
print(summary(pca7, loadings=T))
xyz_long_uni_7 <- xyz_long_7 %>% mutate(value = predict(pca7,xyz_long_7[,3:5])[,1]) %>% select(-X, -Y, -Z) %>% filter(time_series_id==7 | time_series_id==12) %>% mutate(time_series_id = as_factor(time_series_id))
ggplot(xyz_long_uni_7) + geom_line(aes(x=time_index, y=value, colour=time_series_id)) + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("Time Series from Class 7")

xyz_long_8 <- xyz_long %>% filter(class==8)
pca8 <- princomp(xyz_long_8[,3:5], cor=T) #scale differences T
print(summary(pca8, loadings=T))
xyz_long_uni_8 <- xyz_long_8 %>% mutate(value = predict(pca8,xyz_long_8[,3:5])[,1]) %>% select(-X, -Y, -Z) %>% filter(time_series_id==6 | time_series_id==21) %>% mutate(time_series_id = as_factor(time_series_id))
ggplot(xyz_long_uni_8) + geom_line(aes(x=time_index, y=value, colour=time_series_id)) + xlab("Time Index") + ylab("Value") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("Time Series from Class 8")

```



#### D. Multidimensional Scaling

After implementing multidimensional scaling, the corresponding points to each gesture is plotted on the graph. This does not create a useful way to separate gestures. For example, even though gestures 1 and 5 are distinguishable, gestures 1 and 6 are almost indistinguishable.

```{r visualize-mds}
data_binded_columns<-cbind(x_train,y_train[,-1],z_train[,-1])
data_binded_columns[,-1]=scale(data_binded_columns[,-1])
distance = as.matrix(dist(data_binded_columns[,-1], method = "euclidean"))
fit <- cmdscale(distance, k=2) #choose dimension as 2
fit_data <- data_binded_columns %>% select(V1) %>% mutate(coord1=fit[,1], coord2=fit[,2], Classes = as_factor(V1))

ggplot(fit_data) + geom_point(aes(x=coord1, y=coord2, colour=Classes)) + xlab("Coordinate 1") + ylab("Coordinate 2") + theme(plot.title=element_text(hjust = 0.5)) + ggtitle("MDS") 

```


### Reference

[1] J. Liu, Z. Wang, L. Zhong, J. Wickramasuriya, and V. Vasudevan. uWave: Accelerometer-based personalized gesture recognition and its applications. Pervasive Computing and Communications, IEEE International Conference on, 0:1-9, 2009. (link: https://www.recg.org/publications/liu09percom.pdf)
